---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
  [![R-CMD-check](https://github.com/rstrotmann/nif/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rstrotmann/nif/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->
  

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

# NIF

This is a package to create NONMEM Input Format (NIF) data tables from
SDTM-formatted clinical study data.

The [NONMEM](https://www.iconplc.com/solutions/technologies/nonmem) software
that is often used for population pharmacokinetic/pharmacodynamic (PK/PD)
modeling expects the input data set to follow specific conventions summarized in 
*Bauer, R.J. (2019), NONMEM Tutorial Part I: Description of Commands and
Options, With Simple Examples of Population Analysis. CPT Pharmacometrics Syst.
Pharmacol., 8: 525-537* <doi:10.1002/psp4.12404>.

This package provides functions to sequentially aggregate drug administrations,
PK/PD observations and covariates into a NONMEM-compliant analysis data set
(NIF data set), and tools to explore and visualize NIF data sets.

## Installation

You can install the development version of `nif` like this:

``` r
devtools::install_github("rstrotmann/nif", build_vignettes=TRUE)
```

## Example

### Generate a NIF data set

This is a very basic example using sample SDTM data from a fictional single
ascending dose study to create a NIF data set using `make_nif()`:

```{r message = FALSE, warning = FALSE}
library(nif)
library(tidyverse)

sdtm <- examplinib_sad

nif <- nif() %>%
  add_administration(sdtm, "EXAMPLINIB", analyte = "RS2023") %>%
  add_observation(sdtm, "pc", "RS2023", analyte = "RS2023")

head(nif)
```

In many cases, you may want to add further covariates, e.g., baseline creatinine
from the LB domain:

```{r}
nif <- nif %>%
  mutate(COHORT = ACTARMCD) %>%
  add_baseline(sdtm, "lb", "CREAT") %>%
  add_bl_crcl()
```

### Data exploration

The `nif` package provides a range of functions to explore and summarize NIF
files:

```{r, message = FALSE, warning = FALSE, fig.width=3, fig.height=3}
summary(nif)

invisible(capture.output(
  summary(nif) %>%
    plot()
))
```

# Further information

For further guidance see the help for individual functions and the [project
website](https://rstrotmann.github.io/nif/) on github pages.

