---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NIF

<!-- badges: start -->
<!-- badges: end -->

This is a package to create NONMEM input files (NIF) from SDTM data.

## Installation

You can install the development version of `nif` like so:

``` r
devtools::install_github("rstrotmann/nif", build_vignettes=TRUE)
```

## Example

### NIF data set creation

This is a very basic example using a sample SDTM object to create a NIF data
set using `make_nif()`:

```{r message = FALSE, warning = FALSE}
library(nif)
library(tidyverse)

sdtm <- examplinib_sad

make_nif(sdtm) %>% 
  compress() %>%
  head()
```

In many cases, you may want to add further covariates after `make_nif()`,
see this slightly more elaborate example:

```{r}
nif <- make_nif(sdtm, silent = TRUE) %>% 
  compress() %>% 
  mutate(COHORT = ACTARMCD) %>% 
  add_bl_lab(examplinib_sad %>% domain("lb"), "CREAT") %>% 
  add_bl_crcl()

head(nif)
```

### Data exploration

`nif` contains functions to explore and summarize NIF files:

```{r, message = FALSE, warning = FALSE, fig.width=4, fig.height=4}
summary(nif)

invisible(capture.output(
  summary(nif) %>%
    plot()
))
```

# Further information

For further guidance see the help for individual functions and the [project
website](https://rstrotmann.github.io/nif/) on github pages.

