---
title: "Example: NIF data for a multiple-dose study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiple-dose-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F}
library(tidyr)
library(ggplot2)
library(dplyr)
library(stringr)
library(knitr)
library(nif)
```

# OVERVIEW

This vignette walks through the creation of a NIF data set for a prototypical
multiple dose study (study 'RS2023-0022'), as well as some basic exploratory
analyses.

The (entirely fictional) raw study data are provided in SDTM format as
part of the NIF package as 'examplinib_poc'. Actual SDTM data can be loaded 
using `read_sdtm_sas()` of `read_sdtm_xpt()` - for further information see the
documentation for the respective functions.

## Study design

Study 'RS2023-0022' is a single-arm study in which subjects received multiple
doses of 'examplinib' (substance code 'RS2023').
The treatment duration is different across subjects. PK sampling was on Days 1
and 8 of the treatment period. The PK sampling schedule was rich in the initial
subset of subjects and sparse in the others.

## Study SDTM data

SDTM data objects as defined in the NIF package are aggregates of the original
SDTM domains together with some metadata. Let's print the 'examplinib_poc' SDTM
object for a high-level overview on the data:


```{r, echo=T}
examplinib_poc
```

Note that in the EX domain, the administered drug is given as 'EXAMPLINIB' while
in PC, the corresponding analyte is 'RS2023'. The other analyte, 'RS2023487A' is
a metabolite.

To clearly specify these correlations, the SDTM object contains the fields
`examplinib_poc$analyte_mapping` and `examplinib_poc$metabolite_mapping`. Both
need to be specified when the SDTM data set is created. See the documentation to
`add_analyte_mapping()` and `add_metabolite_mapping()` for details.

For this SDTM object, both have been added already.

# CREATE THE NIF DATA SET

Creation of the NIF data set for this study is straightforward. No imputations
are needed, and no study-specific fields are created:

```{r}
nif_poc <- make_nif(examplinib_poc, spec="PLASMA")
```

Of note, the LB domain within this SDTM object contains baseline creatinine data
which can be used to calculate the baseline creatinine clearance as an estimate
for glomerular filtration rate (eGFR). Please see the below code for details.

In the first step, baseline creatinine is added from 'LB', and then the
creatinineclearance is calculated using further demographic parameters (sex,
age, race, weight). The standard method is Cockcroft-Gault but other methods can
be specified.
 
In the last step, the NIF data set, is cleaned up, keeping only the fields
that are commonly used:

```{r}
nif_poc <- make_nif(examplinib_poc, spec="PLASMA") %>% 
  add_bl_lab(examplinib_poc$domains[["lb"]], "CREAT", "SERUM") %>%
  add_bl_crcl() %>% 
  compress_nif()
```

(In fact, the resulting NIF data set, including baseline creatinine clearance is
already included in the package under the name 'examplinib_poc_nif'.)

In case NIF data sets have been generated independently, they can be converted
into NIF objects using `new_nif()`, see the documentation to this
function for details.

The resulting NIF object is essentially wrapper around a data frame which has
the set of variables expected by NONMEM or other population analysis tools. This
becomes visible when the object is converted into a data frame:

```{r, comment="# "}
nif_poc %>% 
  as.data.frame() %>% 
  head()
```

When the NIF object itself is printed, a summary of the data is shown instead:

```{r, comment="# "}
nif_poc
```

# EXPLORATION

## Demographics

A good initial overview on the distribution of baseline parameters, the
administered drugs, analytes, observations, etc. can be inspected with
`summary()`. Since we have above added baseline eGFR to the data set, the output
also summarizes the number of patients with normal renal function or impaired
renal function:

```{r, comment="# "}
summary(nif_poc)
```

For a visual overview of the NIF data set `plot()` can be applied to the summary:

```{r, results='hide', fig.width=3, fig.height=3}
plot(summary(nif_poc))
```

## Exposure

In this study, all `r nrow(subjects(nif_poc))` subject received the same dose level: 

```{r}
nif_poc %>%
  dose_levels() %>% 
  kable(caption="Dose levels")
```

However, there were subjects with dose reductions, as we can see when filtering
the NIF data set for EVID==1 (administrations) and summarizing the administered
dose:

```{r}
nif_poc %>% 
  as.data.frame() %>% 
  filter(EVID==1) %>% 
  group_by(DOSE) %>% 
  summarize(n=n()) %>% 
  kable()
```

Note that in the above code, the nif dataset was first converted into a data
frame as the tidyverse verbs ('filter', 'group_by', 'summarize') are not defined
for NIF objects. This is a common pattern that will often be used for more 
complex analyses!

To identify the subjects with dose reductions: 

```{r, comment="# "}
nif_poc %>%
  dose_red_sbs()
```

Let's have a plot of the doses over time in these subjects: 

```{r fig.width=5, fig.height=2}
nif_poc %>% 
  as.data.frame() %>% 
  filter(ID %in% (nif_poc %>% dose_red_sbs())) %>% 
  filter(EVID==1) %>% 
  ggplot(aes(x=TIME, y=DOSE, color=as.factor(ID))) + 
  geom_point() +
  geom_line() +
  theme(legend.position="none") +
  theme_bw()
```

We see that dose reductions happended at different times during treatment. Another
way of visualizing this is per the `mean_dose_plot()` function:

```{r fig.width=5, fig.height=3}
nif_poc %>%
  mean_dose_plot()
```

The upper panel shows the mean dose over time, and we can see that after ~Day 13,
the mean dose across all treated subjects drops due to dose reductions in some
subjects. To put this into context, the lower panel shows the number of subjects
on treatment over time, and we see that most subjects had treatment durations
of around 30 days. Note the fluctuations that indicate single missed doses in
individual subjects!

## PK sampling

The PK sampling time points in this study were:

```{r}
nif_poc %>% 
  as.data.frame() %>% 
  filter(EVID==0) %>% 
  group_by(NTIME, ANALYTE) %>% 
  summarize(n=n(), .groups="drop") %>% 
  pivot_wider(names_from="ANALYTE", values_from="n") %>% 
  kable(caption="Observations by time point and analyte")
```

From the different numbers of samplings per nominal time point, we see that only
a subset of subjects had a rich sampling scheme. Let's identify those:

```{r, comment="# "}
nif_poc %>% 
  rich_sampling_sbs(analyte="RS2023", max_time=24, n=4)
```

In this code, the function `rich_sampling_sbs()` receives as input the analyte
of interest, the time interval across which the number of sanmples is evaluated,
and the minimum number of samples to qualify the schedule as rich.

## Plasma concentration data

Let's plot the individual and mean plasma concentration profiles on Day 1 for
the parent, RS2023, and the metabolite, RS2023487A:

```{r fig.width=5, fig.height=3}
temp <- nif_poc %>% 
  filter(ID %in% (nif_poc %>%
                    rich_sampling_sbs(analyte="RS2023", n=4, max_time=24)))
temp %>% plot(points=T)
```

For the sampling on Day 1 only:

```{r fig.width=5, fig.height=3, warning=F, message=F}
temp <- temp %>% 
    filter(TIME<=24)

temp %>% plot(analyte="RS2023", points=T, administrations=T, max_x=15)

temp %>% plot(analyte="RS2023", mean=T, y_scale="log")

nif_poc %>% 
  filter(ID %in% (nif_poc %>%
                    rich_sampling_sbs(analyte="RS2023", n=4, max_time=24))) %>% 
  plot(nominal_time=F, min_x=192)
```

The above code showcases some options for the `plot()` function: 

* In addition to the line graphs, the individual data points can be shown with
  the option `points=TRUE`.
* `min_x` and `max_x` can be used to define a speficic time interval.
* In the first 2 graphs, the analyte (RS2023) is specifically selected. In the
  third graph, no analyte is specified, and both analytes are plotted.
* `nonminal_time=TRUE` selects the nominal time for the x axis. The default for
  individual curves is `nominal_time=FALSE`, i.e., the actual time is used. Note
  that for mean profiles as shown in plot 2, nominal time is used automatically
  for the averaging.
* The option `y_scale="log"` can be used to plot the data on a log concentration
  scale.


