---
title: "NIF data for a multiple-dose study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NIF data for a multiple-dose study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup, message=F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(nif)

theme_set(theme_bw())
```

# OVERVIEW

This vignette walks through the creation of a NONMEM input file (NIF) data set
for a prototypical multiple dose study (study 'RS2023-0022'), followed by some
basic exploratory analyses.

The (fictional) raw study data are included as part of the NIF package as
`examplinib_poc`. Custom SDTM data can be loaded using `read_sdtm()`.

## Study design

Study 'RS2023-0022' is a single-arm study in which subjects received multiple
doses of 'examplinib' (substance code 'RS2023').
The treatment duration is different across subjects. PK sampling was on Days 1
and 8 of the treatment period. The PK sampling schedule was rich in the initial
subset of subjects and sparse in the others.

## Study SDTM data

The  package provides the 'sdtm' class as a wrapper to keep all
SDTM domain tables of a clinical study together in one object. The
`examplinib_poc` sdtm object contains the DM, EX, PC and VS domains. Let's
print the `examplinib_poc` sdtm object for a high-level overview:


```{r, echo=T}
examplinib_poc
```

Note that in the EX domain, the administered drug is given as 'EXAMPLINIB' while
in PC, the corresponding analyte is 'RS2023'. The other analyte, 'RS2023487A' is
a metabolite.

To clearly specify these correlations, the SDTM object includes the fields
`examplinib_poc$analyte_mapping` and `examplinib_poc$metabolite_mapping`. Both
can be defined when the SDTM data set is created. See the documentation to
`add_analyte_mapping()` and `add_metabolite_mapping()` for details.

For this SDTM object, both mappings have been added already.

# CREATING A NIF DATA SET

To enable more complex analyses and analyses in the time domain, the data from
different SDTM domains must be integrated into a single data frame in long 
table format. For modeling analyses, a common format specification is the NONMEM
input file (NIF) format that is at the core of the `nif` package.

`nif` objects are minimally composed from administrations and observations.
Following the NONMEM nomenclature, the former have an 'EVID' of 1, the latter an
'EVID' of 0. In the following code, an empty nif object is created using
`new_nif()`, to which 'EXAMPLINIB' administrations are then added using
`add_administration()`. Note that the analyte field for this drug is set to
'RS2023'. In a second step, pharmacokinetic concentrations for 'RS2023' are
added from the 'PC' SDTM domain. To match the administration, the analyte field
is again 'RS2023'. The compartment is automatically set to 2. Both functions
have as their first arguemt the 'sdtm' object:

```{r}
sdtm <- examplinib_poc

nif_poc <- new_nif() %>% 
  add_administration(sdtm, extrt = "EXAMPLINIB", analyte = "RS2023") %>% 
  add_observation(sdtm, domain = "pc", testcd = "RS2023", analyte = "RS2023")
```

The 'LB' domain within this sdtm object contains baseline creatinine data which
can be used to calculate the individual baseline creatinine clearance as an
estimate for glomerular filtration rate (eGFR). Baseline covariates are created
using 'add_baseline()`.

In the first step, baseline creatinine is added from 'LB', and then the
creatinine clearance is calculated using further demographic parameters (sex,
age, race, weight). The standard method is Cockcroft-Gault but other methods can
be specified:

```{r}
nif_poc <- nif_poc %>% 
  add_baseline(sdtm, domain = "lb", testcd = "CREAT") %>% 
  add_bl_crcl()
```

In fact, the nif data set that we have just created, including the baseline
creatinine clearance data, is already included in the `nif` package as
`examplinib_poc_nif`.

In case NIF data sets have been generated externally, they can be converted into
nif objects using `new_nif()`, see the documentation to this function for
details.

Note that a nif object is essentially a wrapper around a data frame with the set
of variables expected by NONMEM or other population analysis tools. This becomes
visible when the object is converted into a data frame:

```{r, comment="# "}
nif_poc %>% 
  as.data.frame() %>% 
  head()
```

When the NIF object itself is printed, a summary of the data is shown instead:

```{r, comment="# "}
nif_poc
```

# EXPLORATION

## Demographics

For an initial overview on the distribution of baseline parameters, the
administered drugs, analytes, observations, etc., NIF objects can be inspected
with the `summary()` function. Note that since we have added baseline eGFR to
the data set, the output also summarizes the number of patients with normal
renal function or impaired renal function:

```{r, comment="# "}
summary(nif_poc)
```

For a visual overview of the NIF data set `plot()` can be applied to the summary:

```{r, results='hide', fig.width=3, fig.height=3}
plot(summary(nif_poc))
```

## Exposure

In this study, all `r nrow(subjects(nif_poc))` subject received the same dose level: 

```{r}
nif_poc %>%
  dose_levels() %>% 
  kable(caption="Dose levels")
```

However, there were subjects with dose reductions, as we can see when filtering
the nif data set for EVID == 1 (administrations) and summarizing the administered
dose:

```{r}
nif_poc %>% 
  filter(EVID==1) %>% 
  group_by(DOSE) %>% 
  summarize(n=n()) %>% 
  kable()
```

To identify the subjects with dose reductions: 

```{r, comment="# "}
nif_poc %>%
  dose_red_sbs()
```

Let's have a plot of the doses over time in these subjects: 

```{r fig.width=5, fig.height=2}
nif_poc %>% 
  filter(ID %in% (nif_poc %>% dose_red_sbs())) %>% 
  filter(EVID == 1) %>% 
  ggplot(aes(x = TIME, y = DOSE, color = as.factor(ID))) + 
  geom_point() +
  geom_line() +
  theme(legend.position="none") 
```

We see that dose reductions happended at different times during treatment. Another
way of visualizing this is per the `mean_dose_plot()` function:

```{r fig.width=5, fig.height=3}
nif_poc %>%
  mean_dose_plot()
```

The upper panel shows the mean dose over time, and we can see that after ~Day 13,
the mean dose across all treated subjects drops due to dose reductions in some
subjects. To put this into context, the lower panel shows the number of subjects
on treatment over time, and we see that most subjects had treatment durations
of around 30 days. Note the fluctuations that indicate single missed doses in
individual subjects!

## PK sampling

The PK sampling time points in this study were:

```{r}
nif_poc %>% 
  filter(EVID == 0) %>% 
  group_by(NTIME, ANALYTE) %>% 
  summarize(n = n(), .groups = "drop") %>% 
  tidyr::pivot_wider(names_from = "ANALYTE", values_from = "n") %>% 
  kable(caption = "Observations by time point and analyte")
```

From the different numbers of samplings per nominal time point, we see that only
a subset of subjects had a rich sampling scheme. Let's identify those:

```{r, comment="# "}
nif_poc %>% 
  rich_sampling_sbs(analyte = "RS2023", max_time = 24, n = 6)
```

In this code, the function `rich_sampling_sbs()` receives as input the analyte
of interest, the time interval across which the number of sanmples is evaluated,
and the minimum number of samples to qualify the schedule as rich.

## Plasma concentration data

Let's plot the individual and mean plasma concentration profiles on Day 1 for
the parent, RS2023, and the metabolite, RS2023487A:

```{r fig.width=5, fig.height=3}
temp <- nif_poc %>% 
  filter(ID %in% (rich_sampling_sbs(nif_poc, analyte = "RS2023", n=4)))
temp %>% plot(dose = 500, points = TRUE, title = "Rich sampling subjects")
```

For single and mutliple dose administrations separately: 

```{r fig.width=5, fig.height=3, warning=F, message=F}
temp <- temp %>% 
  index_rich_sampling_intervals()

temp %>% filter(RICH_N == 1) %>% 
  plot(analyte = "RS2023", mean = TRUE, title = "Single-dose PK")

temp %>% filter(RICH_N == 2) %>% 
  plot(analyte = "RS2023", title = "Multiple-dose PK", time = "NTIME", mean = T)
```

## Non-compartmental analysis

```{r message = F, warning= F}
nca <- examplinib_poc_nif %>% 
  index_rich_sampling_intervals(analyte = "RS2023", min_n = 4) %>% 
  nca("RS2023", group = "RICH_N")

nca %>% 
  nca_summary_table(group = "RICH_N") %>% 
  kable()
```


