---
title: "Example: NIF data for a multiple-dose study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiple-dose-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F}
library(tidyr)
library(ggplot2)
library(dplyr)
library(stringr)
library(knitr)
library(nif)
```

# OVERVIEW

This is an example for a fictional multiple-dose study with a
fictional drug, `RS2023`. As part of the package, a synthetic SDTM data set for
this study (study `RS2023-0022`) is included as `examplinib_poc`:

```{r, echo=F}
print(summary(examplinib_poc))
```

## Study design

This is a single-arm study in which subjects received multiple doses of 'RS2023'. 
The treatment duration is different across subjects. PK sampling was on Days 1
and 8 of the treatment period. The PK sampling schedule was rich in the initial
subset of subjects and sparse in the others.

## Objective

The objective of this vignette is to outline how to approach the data set, 
gradually exploring the subject disposition, treatment durations and the available
pharmacokinetic data. This should always be the first step before entering 
pharmacokinetic modeling.

# CREATE THE NIF DATA SET

Creation of the NIF dataset is straightforward. No imputations are needed, and
no study-specific fields are created:

```{r}
nif_poc <- make_nif(examplinib_poc, spec="PLASMA")
summary(nif_poc)
```

# EXPLORATION

## Demographics

The number of subjects and the sex distribution can already be seen from the
`summary` output (above). We now want to get an overview on the age, body weigth
and body mass index (BMI) distributions, as well as the weigth distribution by
sex and race:

```{r fig.width=3, fig.height=3}
nif_poc %>% age_hist()
nif_poc %>% weight_hist()
nif_poc %>% bmi_hist()
nif_poc %>% wt_by_sex()
nif_poc %>% wt_by_race()
```

## Exposure

In this study, all `r nrow(subjects(nif_poc))` subject received the same dose level: 

```{r}
nif_poc %>%
  dose_levels() %>% 
  kable(caption="Dose levels")
```

However, there were subjects with dose reductions, as we can see when filtering
the NIF data set for EVID==1 (administrations) and summarizing the administered
dose:

```{r}
nif_poc %>% 
  as.data.frame() %>% 
  filter(EVID==1) %>% 
  group_by(DOSE) %>% 
  summarize(n=n()) %>% 
  kable()
```

Note that in the above code, the nif dataset was first converted into a data
frame as the tidyverse verbs ('filter', 'group_by', 'summarize') are not defined
for NIF objects. This is a common pattern that will often be used for more 
complex analyses!

To identify the subjects with dose reductions: 

```{r}
nif_poc %>% dose_red_sbs()
```

Let's have a plot of the doses over time in these subjects: 

```{r fig.width=5, fig.height=2}
nif_poc %>% 
  as.data.frame() %>% 
  filter(ID %in% (nif_poc %>% dose_red_sbs())) %>% 
  filter(EVID==1) %>% 
  ggplot(aes(x=TIME, y=DOSE, color=as.factor(ID))) + 
  geom_point() +
  geom_line() +
  theme(legend.position="none") +
  theme_bw()
```

We see that dose reductions happended at different times during treatment. Another
way of visualizing this is per the `mean_dose_plot()` function:

```{r fig.width=5, fig.height=3}
nif_poc %>% mean_dose_plot()
```

The upper panel shows the mean dose over time, and we can see that after ~Day 13,
the mean dose across all treated subjects drops due to dose reductions in some
subjects. To put this into context, the lower panel shows the number of subjects
on treatment over time, and we see that most subjects had treatment durations
of around 30 days. Note the fluctuations that indicate single missed doses in
individual subjects!

## PK sampling

The PK sampling time points in this study were:

```{r}
nif_poc %>% 
  as.data.frame() %>% 
  filter(EVID==0) %>% 
  group_by(NTIME, ANALYTE) %>% 
  summarize(n=n(), .groups="drop") %>% 
  pivot_wider(names_from="ANALYTE", values_from="n") %>% 
  kable(caption="Observations by time point and analyte")
```

From the different numbers of samplings per nominal time point, we see that only
a subset of subjects had a rich sampling scheme. Let's identify those:

```{r}
nif_poc %>% 
  rich_sampling_sbs(analyte="RS2023", max_time=24, n=4)
```

In this code, the function `rich_sampling_sbs()` receives as input the analyte
of interest, the time interval across which the number of sanmples is evaluated,
and the minimum number of samples to qualify the schedule as rich.

## Plasma concentration data

Let's plot the individual and mean plasma concentration profiles on Day 1 for
the parent, RS2023, and the metabolite, RS2023487A:

```{r fig.width=5, fig.height=3}
temp <- nif_poc %>% 
  filter(ID %in% (nif_poc %>%
                    rich_sampling_sbs(analyte="RS2023", n=4, max_time=24)))
temp %>% plot(points=T)
```

For the sampling on Day 1 only:

```{r fig.width=3, fig.height=3, warning=F, message=F}
temp <- temp %>% 
    filter(TIME<=24)

temp %>% plot(analyte="RS2023", points=T, administrations=T, max_x=15)

temp %>% plot(analyte="RS2023", mean=T, y_scale="log")

nif_poc %>% 
  filter(ID %in% (nif_poc %>%
                    rich_sampling_sbs(analyte="RS2023", n=4, max_time=24))) %>% 
  plot(nominal_time=F, min_x=192)
```

The above code showcases some options for the `plot()` function: 

* In addition to the line graphs, the individual data points can be shown with
  the option `points=TRUE`.
* `min_x` and `max_x` can be used to define a speficic time interval.
* In the first 2 graphs, the analyte (RS2023) is specifically selected. In the
  third graph, no analyte is specified, and both analytes are plotted.
* `nonminal_time=TRUE` selects the nominal time for the x axis. The default for
  individual curves is `nominal_time=FALSE`, i.e., the actual time is used. Note
  that for mean profiles as shown in plot 2, nominal time is used automatically
  for the averaging.
* The option `y_scale="log"` can be used to plot the data on a log concentration
  scale.


