---
title: "Imputations and assumptions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputations and assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This vignette outlines the inner workings of the two main functions provided by
the `nif` package to create nif objects, i.e., `add_administration()` and
`add_obervation()`. It is meant to summarize what assumptions are made and how
missing information is imputed.

# Baseline parameters

The subjects' age is derived from the 'AGE' field in the 'DM' domain (such
domain/field pairs are referred to in the following as, e.g., *DM.AGE*). If
'DM.AGE' is missing, age is derived as the difference between 'DM.RFSTDTC'
and 'DM.BRTHDTC'.

Besides age, the subjects' height and body weight are included as standard
fields in the nif object. Both values are derived from the 'VS' domain, where
the baseline time point is identified as either 'VISIT' = "SCREENING" or
'VSBLFL' = "Y". If multiple measurements fulfilling these condition are found
for a given subject, the mean value is used.

# Administrations

In general, drug administration events are added to a nif object in the
following way:

```{r}
library(dplyr)
library(nif)

my_nif <- new_nif() %>%
  add_administration(examplinib_sad, extrt = "EXAMPLINIB", analyte = "RS2023")
```

The `add_administration()` function uses the drug administration data from 'EX'
with 'EX.EXTRT' filtered for the indicated treatment. If an 'analyte' is
specified, this analyte name is used instead of the 'extrt'.

## Subject filtering

By default, the subjects included by `add_administration()` exclude screening
failures, as well as subjects not treated (see the default 'subject_filter'
string of `!ACTARMCD %in% c('SCRNFAIL', 'NOTTRT')`. Other exclusion filters
can be used, if needed.

## Time imputations

Depending on the study type (single vs. multiple administrations), 'EX' may
define administration episodes spanning multiple administrations, i.e., from
'EX.EXSTDTC' to 'EX.EXENDTC'. A typical example is shown below (some columns
omitted for clarity):

```
STUDYID     USUBJID           EXDOSE  EXDOSFRQ  EXSTDTC           EXENDTC
2023000400  0230004001010002  500     ONCE      2023-08-18T08:06  2023-08-21T08:43
```

`add_administration()` expands administration episodes to individual rows for
each day.

Administration episodes do not include time information for individual days. In
addition, the time part of 'EXSDTDTC' or 'EXENDTC' may be missing. This if often
the case when uncleaned data is analyzed. Since precise time information for 
administration events is essential for all downstream analyses, a series of time
imputations are performed by `add_admininstration()`.

A first set of imputations (see 1. through 3. below) is performed on the
non-expanded EX domain with the aim to ensure that each episode has 'EXSTDTC'
and 'EXENDTC' fields. Imputations 5. and 6. are performed after expansion of the
administration episodes:

### 1. Missing last EXENDTC

If the last administration episode for a subject (and a treatment) has an empty
'EXENDTC', it is replaced with the date/time provided by 'DM.RFENDTC', if
available, i.e., the subject's reference end date. See the documentation for the
(internal) function `impute_exendtc_to_rfendtc()` for further details.

### 2. Ongoing treatment

If after the above imputation attempt, the last administration episode still has
no 'EXENDTC' entry, it is replaced with `cut_off_date`. This situation is
often found in interim analyses where some subjects are still on treatment.
The `cut_off_date` parameter can be specified in the call to
`add_administration()`, or, if not specified, is set to the last administration
event found in the whole dataset (refer to the documentation of
`impute_exendtc_to_cutoff()` for details).

### 3. Missing EXENDTC in other administration episodes

If in an unclean data set, 'EXENDTC' is missing in episodes that are not the
last episode for a given subject and treatment, it is replaced with the day
before the subsequent administration episode start ('EXSTDTC'). It should be
understood that this reflects a rather strong assumption, i.e., that the
treatment was continued into the next administration episode. Consider this a 
last-resort imputation that should be avoided by prior data cleaning, if ever
possible. This imputation, if conducted, therefore issues a warning that cannot
be suppressed with `silent = TRUE` (see the documentation to
`impute_missing_exendtc()` for details).

### 4. Expansion of treatment administration episodes

All administration episodes, i.e., the intervals between 'EXSTDTC' and 'EXENDTC'
for a given row in EX, are expanded into a sequence of rows with one
administration day per row. The administration times for all rows except for
the last are taken from the time information in EXSTDTD, whereas the time
for the last administration event in the respective episode is taken from the
time information in EXENDTC.

### 5. Impute administration time from PCRFTDTC

For administration days for which PK sampling events are recorded in PC, the
administration time is taken from PC.PCRFTDTC, if this field is available. Time
information derived during expansion (see 4.) is overwritten during this
process. See the documentation to `impute_admin_times_from_pcrftdtc()` for
details.

### 6. Carry forward administration time

Finally, for all administration events per subject and treatment, further
missing time information is carried forward from the last available time
information.


<!-- # Observations -->

<!-- ## Specimen -->

<!-- The observations included in the NIF file are the pharmacokinetic concentrations -->
<!-- included in the 'PC' domain. Depending on the design of the underlying clinical -->
<!-- study, there may be data from different specimen types, e.g., blood, plasma, -->
<!-- urine, feces. -->

<!-- The specimen type to be included in the NIF file can be explicitly provided to -->
<!-- `make_nif()`, e.g., `make_nif(sdtm.data, spec="PLASMA")`. If not provided, the -->
<!-- most likely is used, i.e., either "PLASMA" or "BLOOD". -->

<!-- ### Measurements not done -->

<!-- If the *permissible* field `PCSTAT` is used in the PC domain, providing -->
<!-- information on the status of the pharmacokinetic samples, data records with  -->
<!-- `PCSTAT=="NOT DONE"` are discarded. -->

<!-- ### Observation time -->

<!-- The actual absolute date and time for each observation is taken from the `PCDTC` -->
<!-- field. The nominal (planned) time is taken from the `PCELTM` (planned elapsed -->
<!-- time) field. This *permissible* variable has the format "PTxxH" where *xx* is -->
<!-- the time since the last administration. -->

<!-- ### Other fields required by NONMEM -->

<!-- For all observations, the compartment (`CMT`) is fixed to 2. Depending on the -->
<!-- further analysis, e.g., the structural model used in the nonlinear mixed-effects -->
<!-- modeling, the compartment may need to be changed, or different compartments for -->
<!-- the parent compound and metabolites may need to be defined. -->

<!-- The `DV` field in the NIF output is set to `PCSTRESN/1000`. This assumes that -->
<!-- the pharmacokinetic concentrations are given in ng/ml units. As per the NONMEM -->
<!-- conventions, `LNDV` is the natural log of `DV`. `MDV` (*missing dependent -->
<!-- variable*) is set to 1 for entries with missing `DV`. -->

<!-- For all observations, `EVID` is fixed to zero. -->

<!-- ### SEX recoding -->

<!-- As per the SDTM definitions, the subjects' sex is encoded in `SEX` as "M" or -->
<!-- "F". In the NIF output file, "M" is recoded to "0" and "F" to "1" to facilitate -->
<!-- downstream analysis. -->

<!-- ## Administrations -->

<!-- The administrations included in the NIF file come from the EX SDTM domain, but -->
<!-- only administration data from subjects who have corresponding observations are -->
<!-- included. -->

<!-- ### Date expansion -->

<!-- Depending on the study type (single vs. multiple administrations), the EX -->
<!-- domain may define administration intervals, i.e., from `EXSTDTC` to `EXENDTC`.  -->
<!-- A typical example is shown below (some columns omitted for clarity): -->

<!-- ``` -->
<!-- STUDYID     USUBJID           EXDOSE  EXDOSFRQ  EXSTDTC           EXENDTC -->
<!-- 2023000400  0230004001010002  500     ONCE      2023-08-18T08:06  2023-08-21T08:43 -->
<!-- ``` -->
<!-- During NIF generation, `make_nif()` expands the time range to individual -->
<!-- rows for each day. The respective administration time is assumed to be the  -->
<!-- administration time noted for the start date for the whole administration  -->
<!-- interval (i.e., 8:06 in the above example), except for the end date for which -->
<!-- the time specified for that date is taken (i.e., 8:43 in the above example). -->

<!-- ### Analyte mapping -->

<!-- To distinguish administrations and observations relating to multiple drugs,  -->
<!-- `make_nif()` introduces the field `ANALYTE` into the NIF output file. Its -->
<!-- values correspond to the values used in `PCTESTCD` in the PC domain. The SDTM -->
<!-- variable in the EX domain that defines the administered drug is `EXTRT`, and in -->
<!-- many cases, its values do not correspond to the values defined in `PCTESTCD` for -->
<!-- the respective parent compound analytes. In order to associate administrations -->
<!-- to the parent plasma concentrations, mapping information must be provided on  -->
<!-- the level of the `sdtm` object (see `?add_mapping()` for additional -->
<!-- information). -->

<!-- ## Data cut-off and date and time imputations -->

<!-- During the creation of the NIF file, a global cut-off date is defined that -->
<!-- reflects the last observation across all subjects in the PC domain. -->

<!-- In ongoing studies, the treatment end date for some records may not be  -->
<!-- defined, e.g., `EXENDTC` is empty in the respective record in the `EX` domain. -->
<!-- This is often the case when the treatment is still ongoing at the time of the -->
<!-- data cut-off. In these records, `EXENDTC` is set to the cut-off date. -->

<!-- ### Missing treatment start time -->

<!-- Specifically for administrations that are followed by observations on the same -->
<!-- day, the administration time point is critical information to calculate precise -->
<!-- post-dose time points. If time information is not included in EX for such days -->
<!-- (see the example below), `make_nif()` cross references with the `PCRFDTC` -->
<!-- variable in PC, i.e., the reference time recorded the observations on this day. -->

<!-- ``` -->
<!-- STUDYID     USUBJID           EXDOSE  EXDOSFRQ  EXSTDTC     EXENDTC -->
<!-- 2023000400  0230004001010002  500     ONCE      2023-08-18  2023-08-21T08:43 -->
<!-- ``` -->

<!-- That time is then also carried forward to the following administrations in the -->
<!-- whole administration interval. -->

<!-- ### Missing treatment end time -->

<!-- In some cases, no end time is included in `EXENDTC`, see the example below. In  -->
<!-- these cases, `make_nif()` assumes the end time to be the same for `EXSTDTC` and -->
<!-- `EXENDTC`. -->

<!-- ``` -->
<!-- STUDYID     USUBJID           EXDOSE  EXDOSFRQ  EXSTDTC           EXENDTC -->
<!-- 2023000400  0230004001010002  500     ONCE      2023-08-18T08:06  2023-08-21 -->
<!-- ``` -->

<!-- ## TIME -->

<!-- As per the NONMEM requirements, the `TIME` variable in the NIF file gives the -->
<!-- elapsed time in hours since the individual first recorded event, most often the -->
<!-- first pre-dose observation or the first administration. -->





