---
title: "Creating NIF files from SDTM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating NIF files from SDTM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA
)
```

# INTRODUCTION

This is a basic tutorial on using the `nif` package to create NONMEM Input File
(NIF) data sets from Study Data Tabulation Model- (SDTM) formatted data.

## Background

Following current regulatory standards, clinical study data are commonly
provided in SDTM format, an observation-based data tabulation format in which
logically related observations are organized into topical collections (domains).
SDTM is defined and maintained by the [Clinical Data Interchange Standards Consortium (CDISC)](https://www.cdisc.org/standards/foundational/sdtm).

For many exploratory analyses, data from different SDTM domains need to be
combined. For example, demographic and pharmacokinetic concentration data from
the DM and PC domains are both needed to evaluate exposure by age. More complex
analyses like population-level PK and PK/PD analyses may include further data, 
e.g., clinical laboratory, vital sign, or biomarker data.

NONMEM and other commonly used modeling software packages expect the input data
integrated into a (long) data table format with strict requirements to the
formatting and nomenclature of the variables (see, e.g.,
[Bauer, CPT Pharmacometrics Syst. Pharmacol. (2019)](https://doi.org/10.1002/psp4.12404)
for an introduction).
The input data file for these analyses is often casually referred to as a
'NONMEM input file' or 'NIF file', hence the name of this package.

Depending on the nature of the analysis, some of the input variables, may be 
easily and automatically derived from the SDTM-formatted source data, e.g.,
'DOSE' or 'DV' (the dependent variable for pharmacokinetic observations), or
demographic covariates like AGE, SEX or RACE.

Other fields of the input data set
may require study-specific considerations, for example the calculation of
baseline renal or hepatic function categories, definition of specific treatment
conditions by study arm, or the encoding of adverse events or concomitant
medications of interest as categorical covariates.

In contrast to the latter variables which often need manual and study-specific
data programming, the core NIF data set that integrates pharmacokinetic
observations, drug administration information and a standard set of baseline
covariates can often be generated automatically without too much study-specific
user input. It should be however noted that even for such basic NIF files,
missing data points and certain inconsistencies may need to be resolved to get
to an analysis-ready data set, particularly so if the SDTM data stems from an 
ongoing clinical study and has not been fully cleaned yet. More on this point
later.

This package is intended to facilitate the creation of analysis data sets ('NIF
data sets') from SDTM-formatted clinical data. In addition, it includes a set of
functions and tools to explore SDTM and NIF data sets.

The first part of this tutorial describes how to import SDTM data into a `sdtm`
object, and how to explore clinical data on the SDTM level.

The second part of this tutorial walks through the generation of a sample NIF
data set from SDTM data and its initial exploratory analysis.

This tutorial contains live code that depends on the following R packages:

```{r setup, message=F, warning=F}
library(tidyr)
library(dplyr)
library(stringr)
library(nif)
```

# SDTM DATA

## Import SDTM data 

As a minimum, the following SDTM domains are
expected to create a basic NIF file:

* DM - demographic data, mostly age, sex and other variables per subject
* VS - vital signs, particularly weight and height at enrollment, per subject
* EX - exposure information for the administered drugs, e.g., the time of
    administration, dose, etc.
* PC - pharmacokinetic concentrations for the analytes associated with the 
    administered drugs
    
In the most common use case, the respective SDTM data files, e.g., as *.sas7bdat
files, are copied into one folder (in the below example into `path/to/sdtm/data`),
and can then be loaded using:

```{r, eval=FALSE, echo=T}
read_sdtm("path/to/sdtm/data")
```

Based on the above, the minimum required SDTM files are `dm.dat7bdat`,
`vs.dat7bdat`, `ex.dat7bdat` and `pc.dat7bdat`. However, if additional
covariates from AE, LB or any other SDTM domain are to be added, an explicit
list of the domains can be provided to the loading function, e.g., 

```{r, eval=FALSE, echo=T}
read_sdtm("path/to/sdtm/data", domain = c("dm", "vs", "ex", "pc", "lb", "ae"))
```

The return value of this function is a `sdtm` object.

## SDTM objects

`Sdtm` objects are essentially lists of the SDTM domains from a particular
clinical study, plus some metadata. The easiest way of creating `sdtm`
objects is by importing the SDTM data using `read_sdtm_sas()` or
`read_sdtm_xpt()` as shown above. 

The `nif` package includes sample SDTM data sets for demonstration purposes. These
data do not come from actual clinical studies but are fully synthetic data sets
from a fictional single ascending dose (SAD) study (`examplinib_sad`), a
fictional food effect (FE) study (`examplinib_fe`), and a fictional
single-arm proof-of-concept (POC) study with multiple-dose administrations
(`examplinib_poc`).

The original SDTM data can be retrieved from `sdtm` objects by accessing the
individual SDTM domains like demonstrated below for the DM domain of the
`examplinib_fe` data object:

```{r}
examplinib_fe %>%
  domain("dm") %>%
  head(3)
```

The most relevant domains, "dm", "vs", "ex" and "pc", can also be accessed
directly, e.g.:

```{r}
dm <- examplinib_fe$dm
```

Printing an `sdtm` object shows some of its basic properties:

```{r}
examplinib_fe
```

Note the 'Treatment-to-analyte mappings' and 'Parent-to-metabolite mapping'
tables in the above output, we will get back to this when creating a NIF data
set from these SDTM data.

In addition to a general overview, general demographic data for individual 
subjects (by their USUBJID) can be extracted as shown below: 

```{r}
examplinib_fe %>%
  subject_info("20230004001050001")
```

## SDTM suggestions

SDTM data may be incomplete, e.g., when emerging data that have not yet been
fully cleaned are analyzed. In addition, some study-specific
data may be encoded in a non-standardized way, e.g., information on study parts,
cohorts, treatment conditions, etc..

Such data fields may need study-specific considerations and manual imputations
during the creating of the NIF data set. To help deciding which study-specific
factors need to be addressed, the `nif` package includes functions to explore
the structure of SDTM data.

As a starting point, `suggest()` can provide useful suggestions for manual 
imputations for a `sdtm` object:

```{r}
suggest(examplinib_fe)
```

In the case of `examplinib_fe`, there are different treatment arms defined in 
the DM domain that should probably be included as covariates in the NIF file as
they specify the sequence of fasted and fed administrations (suggestion 1).
A respective variable within the NIF data set will be programmed once a basic
NIF file has been generated (see below).

Also, the SDTM data do not provide a relation between treatments and analytes
(suggestion 2). The parent drug name in the SDTM field `EXTRT` (i.e.,
"EXAMPLINIB") is different from the respective analyte for the parent compound
as specified in the field `PCTESTCD` (i.e., "RS2023"). This is a common
situation in data from clinical studies.

To correctly associate
pharmacokinetic concentrations with the administration of the corresponding
drug, a link needs to be provided between the treament and the analyte.
Similarly, metabolites (e.g., "RS2023487A") need to be linked to their parent
compound - particularly if muliple drugs and metabolites are included in the
data set.

### Analyte and metabolite mappings

To establish these relationships, treatment-analyte mappings must be added to 
the `sdtm` object with `add_analyte_mapping()`. In this sample case, the analyte
"RS2023" refers to the (parent) plasma concentration for the drug "EXAMPLINIB." 

To establish the relationships betwen metabolites and their parent copounds, 
`add_metabolite_mapping()` can be used. In the sample data, the `PCTESTCD` of
"RS2023487A" represents a metabolite of "RS2023":

```{r}
sdtm_expl <- examplinib_fe %>% 
  add_analyte_mapping("EXAMPLINIB", "RS2023") %>% 
  add_metabolite_mapping("RS2023", "RS2023487A")
```

# NIF DATA SETS

## Creating NIF data sets

<!-- In addition to re-arranging and joining these SDTM data tables, certain standard -->
<!-- data imputations are automatically performed by the functions provided in this -->
<!-- package, some of which are discussed in more detail in the vignette "Imputations -->
<!-- and assumptions used by make_nif()". -->

The following section summarizes the steps to create a NIF file from an `sdtm`
object.

### Basic NIF file

Most steps in the generation of a basic NIF file from an `sdtm` object are
automated in the `make_nif()` function. The resulting `nif` object is 
a data table with individual rows for individual administrations and
observations and follows the naming conventions summarized in 
[Bauer, CPT Pharmacometrics Syst. Pharmacol. (2019)](https://doi.org/10.1002/psp4.12404).

Standard demographic information including SEX, AGE, RACE is taken from the DM 
domain and merged into the data table. Baseline WEIGHT and HEIGHT are taken from 
the VS domain.

```{r}
nif_expl <- new_nif() %>% 
  add_administration(sdtm_expl, "EXAMPLINIB", analyte = "RS2023") %>% 
  add_observation(sdtm_expl, "pc", "RS2023", cmt = 2)
```

The output of the above code informs that there may be pharmacokinetic data for
multiple specimens in the PC domain, e.g., "PLASMA" and "URINE" or "FECES".
As no specimem was defined, `make_nif()` takes what it considers the the most
likely ("PLASMA" in this case).
Alternatively, the specimen of interest can be explicitly provided
(see `?make_nif()` for details).

Also, a data cut-off was applied that reflects the date and time of the last
observation. This may be relevant in multiple-dose studies where further
administrations may be recorded after the last oberservation. These
administration events are excluded from the NIF file. If this default behavior
is not wanted, the `truncate_to_last_observation` attribute to `make_nif()`
should be set to `FALSE`.

### Study-specific covariates

In many clinical studies, participants are treated in subgroups or serial
treatment periods, for example,
dose escalation cohorts, cross-over periods with different formulations or
co-administration of drug interaction perpetrators, etc.. While DOSE is
automatically included as a continuous covariate, other
categorical covariates may need to be derived, e.g., from the study arm (ARM)
field.

This sample study is a food effect study in which participants received
the test drug, EXAMPLINIB, fasted or fed in a randomized sequence (see `ACTARM`
and `ACTARMCD` in the output of `suggest()`, above). Another field for
consideration may be `EPOCH`:

```{r}
nif_expl <- new_nif() %>% 
  add_administration(sdtm_expl, "EXAMPLINIB", analyte = "RS2023", keep = "EPOCH") %>% 
  add_observation(sdtm_expl, "pc", "RS2023", cmt = 2, keep = "EPOCH")

nif_expl %>% 
  as.data.frame() %>%
  distinct(ACTARMCD, EPOCH)
```

In this example, to create a `FASTED` covariate field, first the period number
is derived from the `EPOCH` field, and then the current `ACTARM` in that
treatment period is decoded to `FASTED`:

```{r}
nif_expl <- nif_expl %>%
  mutate(PERIOD = str_sub(EPOCH, -1, -1)) %>% 
  mutate(TREATMENT = str_sub(ACTARMCD, PERIOD, PERIOD)) %>% 
  mutate(FASTED = case_when(TREATMENT=="A" ~ 1, .default=0)) %>%
  normalize_nif()
```

Note the `silent = TRUE` modifier that omits all messages from `make_nif()`.

As a side note: Alternatively, the SDTM domain YL which provides data on meals 
may be used for the derivation of the food status.

The NIF file includes now the additional fields `PERIOD`, `TREATMENT` and
`FASTED`, see the first few lines of the the data (some fields de-selected to
make the output fit to the page width):

```{r}
nif_expl %>%
  select(USUBJID, PERIOD, TREATMENT, FASTED, TIME, EVID, AMT, ANALYTE, DV) %>% 
  head()
```

## Data disposition

It is generally advisable to explore data sets before proceeding into any more
complex analyses.

The `summary()` function provides a general overview on the data disposition in
a `nif` data set: 

```{r}
summary(nif_expl)
```

Further to the above text output, the `nif` summary object can also be plotted
for a quick graphical overview on the distribution of the baseline demographic
data in the subject population and the analytes included in the `nif` data set.
Note the `ìnvisible(capture.output())` construct around the `plot()` function
in the below code. This is used to omit some non-graphical output:

```{r results='hide', fig.width=3, fig.height=3}
invisible(capture.output(
  nif_expl %>% 
    summary() %>% 
    plot()
))
```

## Plasma concentration data

NIF objects allow basic ploting functionality using `plot()`, directly applied
to the `nif` object. The below shows some examples for common exploratory tasks.

In the simplest exploratory plot, all analytes are shown over time:

```{r, warning=F, fig.width=5, fig.height=4}
plot(nif_expl)
```

To check the integrity of the data set, if often makes sense to plot the analyte
concentrations over time-after-dose (TAD). 

```{r}
nif_expl %>% 
  plot(time = "TAD", analyte = "RS2023", points = T, lines = F, log = T)
```


Modifiers to the `plot()` function allow common adjustments. To show, e.g., the
food effect on the parent compound, "RS2023" in individual subjects, limiting
the time axis to 24 h and showing the nominal time instead of the actual time:

```{r, warning=F, fig.width=5, fig.height=4}
nif_expl %>% 
  plot(analyte = "RS2023", group = "FASTED", max_time = 24, mean = FALSE,
       points = TRUE)
```

The following shows the mean plasma concentration profiles aver all subjects,
over 24 h by the food status:

```{r, warning=F, fig.width=5, fig.height=4}
nif_expl %>% 
  plot( analyte = "RS2023", group = "FASTED", max_time = 24, mean = TRUE)
```

There are further options, refer to `?plot.nif()` for details.

For a quick overview, use `nif_viewer()` to show individual subjects
interactively: 

```{r, eval=FALSE, echo=T}
nif_viewer(nif_expl)
```

## Non-compartmental analysis

The current version of the `nif` package includes rudimentary support for 
non-compartmental pharmacokinetic analysis. the `nca()` function is essentially
a wrapper around the functions provided by the `PKNCA` package.

```{r}
nca <- nif_expl %>% 
  nca(analyte = "RS2023",
      group = "FASTED")

nca %>% 
  nca_summary_table(group="FASTED") %>% 
  kable()
```






