---
title: "Creating NIF files from SDTM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating NIF files from SDTM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA
)
```

# INTRODUCTION

This is a basic tutorial on using the `nif` package to create NONMEM Input File
(NIF) data sets from Study Data Tabulation Model- (SDTM) formatted data.

## Background

Following regulatory standards, clinical study data are commonly provided in
SDTM format, an observation-based data tabulation format in which logically
related observations are organized into topical collections (domains). SDTM is
defined and maintained by the [Clinical Data Interchange Standards Consortium
(CDISC)](https://www.cdisc.org/standards/foundational/sdtm).

To support typical pharmacometric analyses, data from different SDTM domains
need to be first aggregated into a common analysis data set. For example,
demographic and pharmacokinetic concentration data from the DM and PC domains
are both required to evaluate exposure by age. More complex analyses like
population-level PK and PK/PD analyses may include further data, e.g., clinical
laboratory, vital sign, or biomarker data.

NONMEM and other modeling software packages expect the input data provided in
(long) tabular arrangement with strict requirements to the formatting and
nomenclature of the variables (see, e.g., [Bauer, CPT Pharmacometrics Syst.
Pharmacol. (2019)](https://doi.org/10.1002/psp4.12404) for an introduction). The
input data file for these analyses is sometimes casually referred to as a
'NONMEM input file' or 'NIF file', hence the name of this package.

Contingent on the downstream analyses, some of the variables in the analysis
data set can be easily and automatically derived from the SDTM source data,
e.g., 'DOSE' (the administered dose) or 'DV' (the dependent variable for
observations), or demographic covariates like 'AGE', 'SEX' or 'RACE'. Other
fields of the input data set may require study-specific considerations, for
example the calculation of baseline renal or hepatic function categories,
definition of specific treatment conditions by study arm, or the encoding of
adverse events or concomitant medications, as categorical covariates.

While the latter variables often need manual and study-specific data
programming, the core NIF data set can generally be generated automatically
without study-specific user input. It should be noted that even for such basic
NIF files, missing data points and certain inconsistencies may need to be
resolved to get to an analysis-ready data set, particularly so if the SDTM data
stems from an ongoing clinical study and has not been fully cleaned yet. More on
this point later.

This package is intended to facilitate the creation of analysis data sets ('NIF
data sets') from SDTM-formatted clinical data. It also includes a set of
functions and tools to support inital exploration of SDTM and NIF data sets.

## Outline

The [first][SDTM DATA] part of this tutorial describes how to import SDTM data into a `sdtm`
object, and how to explore clinical data on the SDTM level.

The [second][NIF DATA SETS] part of this tutorial walks through the generation of a sample `nif`
data set from SDTM data, and its exploratory analysis.

This tutorial contains live code that depends on the following R packages:

```{r setup, message=F, warning=F}
library(tidyr)
library(dplyr)
library(stringr)
library(nif)
```

# SDTM DATA

## Importing SDTM data 

<!-- As a minimum, the following SDTM domains are -->
<!-- expected to create a basic NIF file: -->

<!-- * DM - demographic data, mostly age, sex and other variables per subject -->
<!-- * VS - vital signs, particularly weight and height at enrollment, per subject -->
<!-- * EX - exposure information for the administered drugs, e.g., the time of -->
<!--     administration, dose, etc. -->
<!-- * PC - pharmacokinetic concentrations for the analytes associated with the  -->
<!--     administered drugs -->
    
In most cases, the source SDTM data are provided as one file per domain, e.g.,
in SAS binary data base storage (.sas7bdat) or SAS Transport File (.xpt) format.

With `path/to/sdtm/data`, the location to the source folder, SDTM data can be
loaded using `read_sdtm()`:

```{r, eval=FALSE, echo=T}
read_sdtm("path/to/sdtm/data")
```

Windows users may want to provide the file path as raw string, i.e., in the form
of

```{r, eval=FALSE, echo=T}
read_sdtm(r"(path\to\sdtm\data)")
```

to ensure that the backslashes in the file path are correctly captured. Note the
parentheses around the file path!

If no domains are explicitly specified, the function attempts to load 'DM',
'VS', 'EX' and 'PC' as a generic set of SDTM domains suitable to create a basic
pharmacokinetic analysis data set.

The return value of this function is a `sdtm` object.

## SDTM objects

`sdtm` objects are essentially aggregates (lists) of the SDTM domains from a
particular clinical study, plus some metadata. The easiest way of creating
`sdtm` objects is by importing the SDTM data using `read_sdtm()` as shown above.

The `nif` package includes sample SDTM data sets for demonstration purposes.
These data do not come from actual clinical studies but are fully synthetic data
sets from a fictional single ascending dose (SAD) study (`examplinib_sad`), a
fictional food effect (FE) study (`examplinib_fe`), and a fictional single-arm
proof-of-concept (POC) study with multiple-dose administrations
(`examplinib_poc`).

The original SDTM data can be retrieved from `sdtm` objects by accessing the
individual SDTM domains like demonstrated below for the DM domain of the
`examplinib_fe` data object:

```{r}
domain(examplinib_sad, "dm") %>% 
  head(3)
```

Printing an `sdtm` object shows relevant summary information:

```{r}
examplinib_fe
```

Note the 'Treatment-to-analyte mappings' and 'Parent-to-metabolite mapping'
tables in the output, we will get back to this when automatically creating NIF
data sets from SDTM data.

Further subject-level summary data can be extracted using `subject_info()`:

```{r}
examplinib_fe %>%
  subject_info("20230004001050001")
```

For a broad-strokes overview on the data disposition, it may be informative to
look at a timeline view of individual domains, e.g. for DM:

```{r fig.width=6, fig.asp = 0.3}
plot(examplinib_fe, "dm")
```

## SDTM suggestions

SDTM data may be incomplete, e.g., when emerging data that have not yet been
fully cleaned are analyzed. In addition, some study-specific
data may be encoded in a non-standardized way, e.g., information on study parts,
cohorts, treatment conditions, etc..

Such data fields may need study-specific considerations and manual imputations
during the creating of the NIF data set. To help deciding which study-specific
factors need to be addressed, the `nif` package includes functions to explore
the structure of SDTM data.

As a starting point, `suggest()` can provide useful suggestions for manual 
imputations for a `sdtm` object:

```{r results="asis"}
suggest(examplinib_fe)
```

In the case of `examplinib_fe`, there are different treatment arms defined in 
the DM domain that should probably be included as covariates in the NIF file as
they specify the sequence of fasted and fed administrations (suggestion 1).
A respective variable within the NIF data set will be programmed once a basic
NIF file has been generated (see below).

Also, the SDTM data do not provide a relation between treatments and analytes
(suggestion 2). The parent drug name in the SDTM field `EXTRT` (i.e.,
"EXAMPLINIB") is different from the respective analyte for the parent compound
as specified in the field `PCTESTCD` (i.e., "RS2023"). This is a common
situation in data from clinical studies.

To correctly associate
pharmacokinetic concentrations with the administration of the corresponding
drug, a link needs to be provided between the treament and the analyte.
Similarly, metabolites (e.g., "RS2023487A") need to be linked to their parent
compound - particularly if muliple drugs and metabolites are included in the
data set.

### Analyte and metabolite mappings

To establish these relationships, treatment-analyte mappings must be added to 
the `sdtm` object with `add_analyte_mapping()`. In this sample case, the analyte
"RS2023" refers to the (parent) plasma concentration for the drug "EXAMPLINIB." 

To establish the relationships betwen metabolites and their parent copounds, 
`add_metabolite_mapping()` can be used. In the sample data, the `PCTESTCD` of
"RS2023487A" represents a metabolite of "RS2023":

```{r}
sdtm_expl <- examplinib_fe %>% 
  add_analyte_mapping("EXAMPLINIB", "RS2023") %>% 
  add_metabolite_mapping("RS2023", "RS2023487A")
```

# NIF DATA SETS

## Creating NIF data sets

<!-- In addition to re-arranging and joining these SDTM data tables, certain standard -->
<!-- data imputations are automatically performed by the functions provided in this -->
<!-- package, some of which are discussed in more detail in the vignette "Imputations -->
<!-- and assumptions used by make_nif()". -->

The following section summarizes the steps to create a NIF file from an `sdtm`
object.

### Basic NIF file

Most steps in the generation of a basic NIF file from an `sdtm` object are
automated in the `make_nif()` function. The resulting `nif` object is 
a data table with individual rows for individual administrations and
observations and follows the naming conventions summarized in 
[Bauer, CPT Pharmacometrics Syst. Pharmacol. (2019)](https://doi.org/10.1002/psp4.12404).

Standard demographic information including SEX, AGE, RACE is taken from the DM 
domain and merged into the data table. Baseline WEIGHT and HEIGHT are taken from 
the VS domain.

```{r}
nif_expl <- new_nif() %>% 
  add_administration(sdtm_expl, "EXAMPLINIB", analyte = "RS2023") %>% 
  add_observation(sdtm_expl, "pc", "RS2023", cmt = 2)
```

The output of the above code informs that there may be pharmacokinetic data for
multiple specimens in the PC domain, e.g., "PLASMA" and "URINE" or "FECES".
As no specimem was defined, `make_nif()` takes what it considers the the most
likely ("PLASMA" in this case).
Alternatively, the specimen of interest can be explicitly provided
(see `?make_nif()` for details).

Also, a data cut-off was applied that reflects the date and time of the last
observation. This may be relevant in multiple-dose studies where further
administrations may be recorded after the last oberservation. These
administration events are excluded from the NIF file. If this default behavior
is not wanted, the `truncate_to_last_observation` attribute to `make_nif()`
should be set to `FALSE`.

### Study-specific covariates

In many clinical studies, participants are treated in subgroups or serial
treatment periods, for example,
dose escalation cohorts, cross-over periods with different formulations or
co-administration of drug interaction perpetrators, etc.. While DOSE is
automatically included as a continuous covariate, other
categorical covariates may need to be derived, e.g., from the study arm (ARM)
field.

This sample study is a food effect study in which participants received
the test drug, EXAMPLINIB, fasted or fed in a randomized sequence (see `ACTARM`
and `ACTARMCD` in the output of `suggest()`, above). Another field for
consideration may be `EPOCH`:

```{r}
nif_expl <- new_nif() %>% 
  add_administration(sdtm_expl, "EXAMPLINIB", analyte = "RS2023", keep = "EPOCH") %>% 
  add_observation(sdtm_expl, "pc", "RS2023", cmt = 2, keep = "EPOCH")

nif_expl %>% 
  as.data.frame() %>%
  distinct(ACTARMCD, EPOCH)
```

In this example, to create a `FASTED` covariate field, first the period number
is derived from the `EPOCH` field, and then the current `ACTARM` in that
treatment period is decoded to `FASTED`:

```{r}
nif_expl <- nif_expl %>%
  mutate(PERIOD = str_sub(EPOCH, -1, -1)) %>% 
  mutate(TREATMENT = str_sub(ACTARMCD, PERIOD, PERIOD)) %>% 
  mutate(FASTED = case_when(TREATMENT=="A" ~ 1, .default=0)) %>%
  normalize_nif()
```

Note the `silent = TRUE` modifier that omits all messages from `make_nif()`.

As a side note: Alternatively, the SDTM domain YL which provides data on meals 
may be used for the derivation of the food status.

The NIF file includes now the additional fields `PERIOD`, `TREATMENT` and
`FASTED`, see the first few lines of the the data (some fields de-selected to
make the output fit to the page width):

```{r}
nif_expl %>%
  select(USUBJID, PERIOD, TREATMENT, FASTED, TIME, EVID, AMT, ANALYTE, DV) %>% 
  head()
```

## Data disposition

It is generally advisable to explore data sets before proceeding into any more
complex analyses.

The `summary()` function provides a general overview on the data disposition in
a `nif` data set: 

```{r}
summary(nif_expl)
```

Further to the above text output, the `nif` summary object can also be plotted
for a quick graphical overview on the distribution of the baseline demographic
data in the subject population and the analytes included in the `nif` data set.
Note the `ìnvisible(capture.output())` construct around the `plot()` function
in the below code. This is used to omit some non-graphical output:

```{r results='hide', fig.width=3, fig.height=3}
invisible(capture.output(
  nif_expl %>% 
    summary() %>% 
    plot()
))
```

## Plasma concentration data

NIF objects allow basic ploting functionality using `plot()`, directly applied
to the `nif` object. The below shows some examples for common exploratory tasks.

In the simplest exploratory plot, all analytes are shown over time:

```{r, warning=F, fig.width=5, fig.height=4}
plot(nif_expl)
```

To check the integrity of the data set, if often makes sense to plot the analyte
concentrations over time-after-dose (TAD). 

```{r}
nif_expl %>% 
  plot(time = "TAD", analyte = "RS2023", points = T, lines = F, log = T)
```


Modifiers to the `plot()` function allow common adjustments. To show, e.g., the
food effect on the parent compound, "RS2023" in individual subjects, limiting
the time axis to 24 h and showing the nominal time instead of the actual time:

```{r, warning=F, fig.width=5, fig.height=4}
nif_expl %>% 
  plot(analyte = "RS2023", color = "FASTED", max_time = 24, mean = FALSE,
       points = TRUE)
```

The following shows the mean plasma concentration profiles aver all subjects,
over 24 h by the food status:

```{r, warning=F, fig.width=5, fig.height=4}
nif_expl %>% 
  plot( analyte = "RS2023", group = "FASTED", max_time = 24, mean = TRUE)
```

There are further options, refer to `?plot.nif()` for details.

For a quick overview, use `nif_viewer()` to show individual subjects
interactively: 

```{r, eval=FALSE, echo=T}
nif_viewer(nif_expl)
```

## Non-compartmental analysis

The current version of the `nif` package includes rudimentary support for 
non-compartmental pharmacokinetic analysis. the `nca()` function is essentially
a wrapper around the functions provided by the `PKNCA` package.

```{r}
nca <- nif_expl %>% 
  nca(analyte = "RS2023",
      group = "FASTED")

nca %>% 
  nca_summary_table(group="FASTED") %>% 
  kable()
```






