---
title: "Making NIF files from SDTM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nif-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a tutorial on using the `nif` package to create NONMEM input files (NIF)
from SDTM data.

```{r echo=F, message=F, warning=F}
library(tidyverse)
```

## Import SDTM data 

The `nif` package is meant to help creating NIF files for pharmacokinetic
(PK) data, e.g., to support population PK modeling.

As a minimum, the following SDTM domains are expected to create a basic NIF file:

* DM - demographic data, mostly age, sex and other variables per subject
* VS - vital signs, particularly weight and height at enrollment, per subject
* EX - exposure information for the administered drugs, e.g., the time of
    administration, dose, etc.
* PC - pharmacokinetic concentrations for the analytes associated with the 
    administered drugs
    
In the most common use case, the respective SDTM data files, e.g., as *.sas7bdat
files, are copied into one folder (in the below example into `path/to/sdtm/data`),
and can then be loaded using:

```{r, eval=FALSE, echo=T}
nif::read_sdtm_sas("path/to/sdtm/data")
```

As said above, the minimum required SDTM files are `dm.dat7bdat`, `vs.dat7bdat`, 
`ex.dat7bdat` and `pc.dat7bdat`. However, if additional covariates from AE, LB
or any other SDTM domain are to be added, an explicit list of the domains can 
be provided to the loading function, e.g., 

```{r, eval=FALSE, echo=T}
read_sdtm_sas("path/to/sdtm/data", "dm", "vs", "ex", "pc", "lb", "ae")
```

The return value of the function is a `sdtm` object. As part of the `nif`
package, sample data for a fictional study with the drug 'examplini' is
included:

```{r}
sdtm.expl <- nif::sdtm(nif::examplinib)
sdtm.expl
```

As shown above, printing an `sdtm` object shows basic information on the data
set.

## Suggestions

Some fields in the SDTM data are study specific, and is therefore hard to
fully automate the generation of standard NIF files for downstream analysis
steps. This needs study-specific considerations and manual adaptations.

As a starting point the function `suggest()` can provide useful suggestions for 
this step:

```{r}
nif::suggest(sdtm.expl)
```

In this case, there are different treatment arms defined in the DM domain that
should probably be included as covariates in the NIF file (see suggestion 1). 
This will be programmed once a basic NIF file has been generated (see below).

Also, the SDTM data do not provide a link between treatments and analytes (see
suggestion 2). This is a common situation that is described in the below
section:

### Analyte mappings

Often, the names for the treatments as listed in the EXTRT field of EX do not
directly correspond to the names of the respective (parent) analytes codes used
in the PC domain (i.e., in the PCTESTCD field).

To establish these relationships, treatment-analyte mappings must be added to 
the `sdtm` object. In this sample case, the analyte 'RS2023' refers to the
(parent) plasma concentration for the drug 'examplinib':

```{r}
sdtm.expl <- nif::sdtm(nif::examplinib) %>% 
  nif::add_mapping("EXAMPLINIB", "RS2023")
```

## Create a NIF file

The following section details how to create NIF files from an `sdtm` object as 
created above, and what assumptions and imputations underlie the algorithm.

### Basic NIF file

In the first step, `make_nif()` is used to create a data table that has
individual rows for each administration and observation. Following the NONMEM
convention for oral administration, the `EVID` field is 1 for the former, and 0
for the latter.

> **NOTE:** If other models are used, ...

Standard demographic information including SEX, AGE, RACE is taken from the DM 
domain and merged into the data table. Baseline WEIGHT and HEIGHT are taken from 
the VS domain.

```{r}
nif.expl <- nif::make_nif(sdtm.expl)
```

The output of the function reminds that there may be pharmacokinetic data fro 
multiple specimens in the PC domain, e.g., PLASMA and URINE or FECES. The exact
specimen can be explicitly provided, e.g., `make_nif(sdtm.expl, spec="PASMA)`.








