---
title: "Creating NIF files from SDTM data"
author: "Rainer Strotmann"
date: "17-Sep-2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating NIF files from SDTM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a basic tutorial on using the `nif` package to create NONMEM Input Files
(NIF) from Study Data Tabulation Model (SDTM) data.

The package is intended to facilitate providing the required data format for
population PK or PK/PD modeling. The source data are typically contained in 
different SDTM tables and need to be joined together, i.e., including but not
limited to baseline demographic and vital sign data, drug exposure data, and
pharmacokinetic concentrations.

In addition to re-arranging and joining the data, certain standard data
imputations are automatically performed by the functions provided in this
package, some of which are discussed in more detail in the vignette "Imputations
and assumptions used by make_nif()".

This tutorial contains live code that depends on the following R packages:

```{r setup, message=F, warning=F}
library(tidyr)
library(dplyr)
library(stringr)
library(knitr)
library(nif)
```

## Import SDTM data 

As a minimum, the following SDTM domains are
expected to create a basic NIF file:

* DM - demographic data, mostly age, sex and other variables per subject
* VS - vital signs, particularly weight and height at enrollment, per subject
* EX - exposure information for the administered drugs, e.g., the time of
    administration, dose, etc.
* PC - pharmacokinetic concentrations for the analytes associated with the 
    administered drugs
    
In the most common use case, the respective SDTM data files, e.g., as *.sas7bdat
files, are copied into one folder (in the below example into `path/to/sdtm/data`),
and can then be loaded using:

```{r, eval=FALSE, echo=T}
read_sdtm_sas("path/to/sdtm/data")
```

Based on the above, the minimum required SDTM files are `dm.dat7bdat`,
`vs.dat7bdat`, `ex.dat7bdat` and `pc.dat7bdat`. However, if additional
covariates from AE, LB or any other SDTM domain are to be added, an explicit
list of the domains can be provided to the loading function, e.g., 

```{r, eval=FALSE, echo=T}
read_sdtm_sas("path/to/sdtm/data", "dm", "vs", "ex", "pc", "lb", "ae")
```

The return value of the function is a `sdtm` object.

## SDTM objects

`sdtm` objects are essentially lists of the SDTM domains from a particular
clinical study. The most common way of creating `sdtm` objects is by importing
the SDTM data using `read_sdtm_sas()` or `read_sdtm_xpt()` (see above). 

The original data can be retrieved accessing the individual domains like
demonstrated below for the (fictional) sample data set `examplinib_fe`:

```{r}
examplinib_fe$domains[["dm"]] %>% head()
```

The most relevant domains, "dm", "vs", "ex" and "pc", can also be accessed
directly, e.g.:

```{r}
dm <- examplinib_fe$dm
```

Printing an `sdtm` object shows basic information on the data set:

```{r}
examplinib_fe
```


## Suggestions

Some fields in the original SDTM data are study specific, and is therefore hard
to fully automate the generation of standard NIF files for downstream analysis
steps. This needs study-specific considerations and manual adaptations.

As a starting point the function `suggest()` can provide useful suggestions for 
a `sdtm`object:

```{r}
suggest(examplinib_fe)
```

In the case of `examplinib_fe`, there are different treatment arms defined in the
DM domain that should probably be included as covariates in the NIF file (see
suggestion 1). This will be programmed once a basic NIF file has been generated
(see below).

Also, the SDTM data do not provide a relation between treatments and analytes
(see suggestion 2). The parent drug name in the SDTM field `EXTRT` (i.e.,
"EXAMPLINIB") is different from the respective analyte for the parent compound as
specified in the field `PCTESTCD` (i.e., "RS2023"). This is a common situation
in data from clinical studies. However, to correctly associate pharmacokinetic
concentrations with the administration of the corresponding drug, a link needs
to be provided between the treament and the analyte. Similarly, metabolites
(e.g., "RS2023487A") need to be linked to their parent compound - particularly
if muliple drugs and metabolites are included in the data set.

Both are described below.


### Analyte mappings

Often, the names for the treatments as listed in the `EXTRT` field of EX do not
directly correspond to the names of the respective (parent) analytes codes used
in the PC domain (i.e., in the `PCTESTCD` field).

To establish these relationships, treatment-analyte mappings must be added to 
the `sdtm` object with `add_analyte_mapping()`. In this sample case, the analyte
"RS2023" refers to the (parent) plasma concentration for the drug "EXAMPLINIB." 

To establish the relationships betwen metabolites and their parent copounds, 
`add_metabolite_mapping()` can be used. In the sample data, the `PCTESTCD` of
"RS2023487A" represents a metabolite of "RS2023":

```{r}
sdtm_expl <- examplinib_fe %>% 
  add_analyte_mapping("EXAMPLINIB", "RS2023") %>% 
  add_metabolite_mapping("RS2023", "RS2023487A")
```

## Creating NIF files

The following section details how to create NIF files from an `sdtm` object.

### Basic NIF file

Most steps in the generation of a basic NIF file from an `sdtm` object are
automated in the `make_nif()` function. The resulting `nif` object is essentially
a data table that has individual rows for each administration and observation.
Following the NONMEM convention for oral administration, the `EVID` field is "1"
for the former, and "0" for the latter.

Standard demographic information including SEX, AGE, RACE is taken from the DM 
domain and merged into the data table. Baseline WEIGHT and HEIGHT are taken from 
the VS domain.

```{r}
nif_expl <- make_nif(sdtm_expl)
```

The output of the above code informs that there may be pharmacokinetic data for
multiple specimens in the PC domain, e.g., "PLASMA" and "URINE" or "FECES".
As no specimem was defined, `make_nif()` takes the most likely ("PLASMA").
Alternatively, the specimen of interest can be explicitly provided
(see `?make_nif()` for details).

Also, a data cut-off was applied that reflects the date and time of the last
observation. This may be most relevant in multiple-dose studies where further
administrations may be recorded after the last oberservation. These
administration events are excluded from the NIF file.

### Study-specific covariates

In many clinical studies, participants are treated in subgroups, e.g.,
dose groups, formulations, co-administered drug interaction
perpetrators. While dose is automatically included as a continuous covariate per
the EX domain, other categorical covariates may need to be derived, e.g., from the
study arm field.

The inclcuded sample study is a food effect study in which participants received
the test drug, EXAMPLINIB, fasted or fed in a randomized sequence (see `ACTARM`
and `ACTARMCD` in the output of `suggest()` above). Often, relevant fields
for consideration are `EPOCH` and `ACTARMCD`:

```{r}
nif_expl %>% 
  as.data.frame() %>% 
  distinct(EPOCH, ACTARMCD, ACTARM)
```

To create a `FASTED` covariate for the sample data set, first the period number
is derived from the `EPOCH` field, and then the current `ACTARM` in that
treatment period is decoded to `FASTED`:

```{r}
nif_expl <- make_nif(sdtm_expl, silent=T) %>%
  mutate(PERIOD=str_sub(EPOCH, -1, -1)) %>% 
  mutate(TREATMENT=str_sub(ACTARMCD, PERIOD, PERIOD)) %>% 
  mutate(FASTED=case_when(TREATMENT=="A" ~ 1, .default=0))
```

Note the `silent=T` modifier that omits all messages from `make_nif()`.

The NIF file has now the additional fields `PERIOD`, `TREATMENT` and `FASTED`,
see the first few lines of the the data:

```{r}
nif_expl %>%
  select(USUBJID, PERIOD, TREATMENT, FASTED, TIME, EVID, AMT, ANALYTE, DV) %>% 
  arrange(USUBJID, PERIOD, TREATMENT, FASTED, TIME, EVID, ANALYTE) %>% 
  as.data.frame() %>% 
  head() %>% 
  kable()
```

## Graphical exploration

NIF objects allow basic ploting functionality using `plot()`, directly applied
to the object. The below shows some examples for common exploratory tasks (for 
more detail, refer to `?plot.nif()`).

Obviously, more complex custom plots (or other analyses) can be done manually on
the `nif` data. Note that the `nif` object must be converted to a data frame for
this purpose (i.e., using `as.data.frame()`):

In the simplest exploratory plot, all analytes are shown:

```{r, warning=F, fig.width=4, fig.height=3}
plot(nif_expl)
```

Modifiers to the `plot()` function allow common adjustments. To show, e.g., the
food effect on the parent compound, "RS2023" in individual subjects, limiting
the time axis to 24 h and showing the nominal time instead of the actual time:

```{r, warning=F, fig.width=4, fig.height=3}
plot(nif_expl, analyte="RS2023", group="FASTED", max_x=24, mean=F, points=T, nominal_time=T)
```

The following shows the mean plasma concentration profiles over 24 h by food
status:

```{r, warning=F, fig.width=4, fig.height=3}
plot(nif_expl, analyte="RS2023", group="FASTED", max_x=24, mean=T)
```

There are further display options, refer to `?plot.nif()` for details.

For a quick overview, use `nif_viewer()` to show individual subjects
interactively: 

```{r, eval=FALSE, echo=T}
nif_viewer(nif_expl)
```

