---
title: "Creating NIF files from SDTM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating NIF files from SDTM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  # comment = "#>"
  comment = "#>"
)
```

# INTRODUCTION

This is a basic tutorial on using the `nif` package to create NONMEM Input
Format (NIF) data sets from Study Data Tabulation Model (SDTM) formatted data.

## Background

Following regulatory standards, clinical study data are commonly provided in
SDTM format, an observation-based data tabulation format in which logically
related observations are organized into topical collections (domains). SDTM is
defined and maintained by the [Clinical Data Interchange Standards Consortium
(CDISC)](https://www.cdisc.org/standards/foundational/sdtm).

To support typical pharmacometric analyses, data from different SDTM domains
need to be aggregated into a single analysis data set. For example,
demographic and pharmacokinetic concentration data from the DM and PC domains
are both required to evaluate exposure by age. More complex analyses like
population-level PK and PK/PD analyses may include further data, e.g., clinical
laboratory, vital sign, or biomarker data.

NONMEM and other modeling software packages expect the input data provided in
(long) tabular arrangement with strict requirements to the formatting and
nomenclature of the variables (see, e.g., [Bauer, CPT Pharmacometrics Syst.
Pharmacol. (2019)](https://doi.org/10.1002/psp4.12404) for an introduction). The
input data file for these analyses is sometimes casually referred to as a
'NONMEM input format' or 'NIF' file, hence the name of this package.

Contingent on the downstream analyses, some of the variables in the analysis
data set can be easily and automatically derived from the SDTM source data,
e.g., 'DOSE' (the administered dose) or 'DV' (the dependent variable for
observations), or demographic covariates like 'AGE', 'SEX' or 'RACE'. Other
fields of the input data set may require study-specific considerations, for
example the calculation of baseline renal or hepatic function categories,
definition of specific treatment conditions by study arm, or the encoding of
adverse events or concomitant medications, as categorical covariates.

While the latter variables often need manual and study-specific data
programming, the core NIF data set can in most cases be generated in a quite
standardized way. But also the manual addition of further analysis-specific
fields is substantially simplified by the functions that the `nif` package
provides. Often, analysis data sets can be thus be created with only a handful
of lines of code.

Even for basic NIF data sets, missing data or data inconsistencies, specifically
those relating to date and time fields, need to be resolved to get to
analysis-ready data sets. This is frequently encountered when analyzing
preliminary data from ongoing clinical studies that have not been fully cleaned.
The `nif` package provides a number of standardized imputation rules to resolve
these issues. More on this point later as well in a separate vignette
(`vignette("nif-imputations")`).

Overall, the `nif` package is intended to facilitate the creation of analysis
data sets ('NIF data sets') from SDTM-formatted clinical data.

## Outline

The [first][SDTM DATA] part of this tutorial describes how to import SDTM data
into a `sdtm` object, and how to explore clinical data on the SDTM level.

The [second][NIF DATA SETS] part walks through the generation of a sample `nif`
data set from SDTM data to illustrate the general workflow for building analysis
data sets.

Finally, the [third][DATA EXPLORATION] part showcases some functions to quickly
explore analysis data sets.

This tutorial contains live code that depends on the following R packages:

```{r setup, message=F, warning=F}
library(tidyr)
library(dplyr)
library(stringr)
library(nif)
```

# SDTM DATA

## Importing SDTM data 
    
In many cases, source SDTM data are provided as one file per domain, often in
SAS binary data base storage format (.sas7bdat) or SAS Transport File (.xpt)
format.

The SDTM data can be loaded using `read_sdtm()`:

```{r, eval=FALSE, echo=T}
read_sdtm("path/to/sdtm/data")
```

Windows users may want to provide the file path as raw string, i.e., in the form
of:

```{r, eval=FALSE, echo=T}
read_sdtm(r"(path\to\sdtm\data)")
```

This ensures that the backslashes in the file path are correctly captured. Note
the inner parentheses around the file path!

If no domains are explicitly specified, all files with the respective extension
are loaded (excluding file names that start with an underscore).

The return value of this function is a `sdtm` object.

## SDTM objects

`sdtm` objects are essentially aggregates (lists) of the SDTM domains from a
particular clinical study, plus some metadata. The easiest way of creating
`sdtm` objects is by importing the SDTM data using `read_sdtm()` as shown above.

The `nif` package includes sample SDTM data sets for demonstration purposes.
These data do not come from actual clinical studies but are fully synthetic data
sets from a fictional single ascending dose (SAD) study (`examplinib_sad`), a
fictional food effect (FE) study (`examplinib_fe`), and a fictional single-arm
proof-of-concept (POC) study with multiple-dose administrations
(`examplinib_poc`).

Printing an `sdtm` object shows relevant summary information:

```{r}
examplinib_fe
```

High-level subject-level disposition data for a specific subject can be
extracted using `subject_info()`:

```{r}
examplinib_fe %>%
  subject_info("20230004001050001")
```

The original SDTM data for specific domains can be retrieved from `sdtm` objects
with the `domain()` function:

```{r}
domain(examplinib_sad, "dm") %>% 
  head(3)
```

`summary()` generates a high-level data disposition overview on a specific
domain:

```{r}
summary(domain(examplinib_sad, "pc"), silent = TRUE)
```

A graphical timeline overview can be generated for a specific domain by, e.g.:

```{r fig.width=7, fig.asp = 0.3}
plot(examplinib_fe, "dm")
```

## SDTM suggestions

SDTM data may be incomplete, e.g., when emerging data that have not yet been
fully cleaned are analyzed. In addition, some study-specific data may be encoded
in a non-standardized way, e.g., information on study parts, cohorts, treatment
conditions, etc..

Such data fields may need study-specific considerations and manual imputations
during the creating of the analysis data set. To help deciding which
study-specific factors need to be addressed, the `nif` package includes
functions to explore the structure of SDTM data.

As we will see in [NIF DATA SETS][NIF DATA SETS], nif objects are built
sequentially from the SDTM data set. Based on the actual data disposition, the
`suggest()` function provide useful code snippets for the creation of analysis
data sets:

```{r results="asis"}
suggest(examplinib_fe)
```

In the above example, suggestions 1 and 2 provide code for the creation of
administration and observation events. We will use this code directly as
provided in section [Creating NIF data sets].

Suggestion 3 notes that the DM domain defines different treatment arms that
should probably be included as covariates in the analysis data set. They
specify the sequence of fasted and fed administrations in this study. We will
deal with this in [Study-specific covariates].

# NIF DATA SETS

The following sections continue using the `examplinib_fe` example to demonstrate
how a `nif` object is created from the `sdtm` data object.

## Basic NIF file

Depending on the analysis needs, `nif` objects are built in a stepwise manner,
starting from an empty `nif` object, sequentially adding treatment
administrations, observations, and covariate fields. The result is a data table
with individual rows for administrations and observations that follows the
naming conventions summarized in [Bauer, CPT Pharmacometrics Syst. Pharmacol.
(2019)](https://doi.org/10.1002/psp4.12404).

The basic `nif` object automatically includes standard demographic parameters as
subject-level covariates: SEX, AGE and RACE, and baseline WEIGHT and HEIGHT are
taken from the DM and VS domains, respectively, and merged into the data set as
columns of those names:

```{r message = FALSE}
sdtm <- examplinib_fe

nif <- new_nif() %>% 
  add_administration(sdtm, "EXAMPLINIB", analyte = "RS2023") %>% 
  add_observation(sdtm, "pc", "RS2023")
```

Note that in the above, the name of the treatment, i.e., the value of the
'EXTRT' field is 'EXAMPLINIB' while the pharmacokinetic analyte name
('PCTESTCD') is 'RS2023'. To harmonize both, the 'analyte' parameter in
`add_administration()` was set to 'RS2023', too.

These are the first rows of the resulting data table:

```{r}
head(nif, 5)
```

## Multiple analytes

To demonstrate how to add multiple analytes to a `nif` object, we will
temporarily switch to another built-in sample data set, `examplinib_sad`. This
`sdtm` object includes pharmacokinetic concentration data for the M1 metabolite
of 'EXAMPLINIB' under the PCTESTCD of 'RS2023487A'. Note how in the below code,
the two analyte observations are added to the data set, setting the metabolite's
name to 'M1', and how its relation to the parent compound is established using
the 'parent' parameter:

```{r message = FALSE}
sdtm1 <- examplinib_sad

nif1 <- new_nif() %>% 
  add_administration(sdtm, "EXAMPLINIB", analyte = "RS2023") %>% 
  add_observation(sdtm, "pc", "RS2023") %>% 
  add_observation(sdtm, "pc", "RS2023487A", analyte = "M1", parent = "RS2023")
```

In analogy to PK observations, observations from any SDTM domain, e.g., LB, VS,
MB, TR, etc., can be added in very much the same way. Please see the
documentation to `add_observation()` for details. This is a powerful feature
that allows effortless construction of analysis data sets for population PK/PD
modeling.

## Study-specific covariates

In this simulated study, participants had received the test drug, EXAMPLINIB,
under fasted or fed conditions in a randomized sequence (see 'ACTARM' and
'ACTARMCD' in the output of `suggest()`), where the 'EPOCH' field in the EX
domain provides information on the current treatment period. It should be noted
that the way such information is encoded in the SDTM data varies considerably.
This is an example, and the specifics of how covariate information can extracted
from a SDTM data set will differ across studies. However, `nif` objects are
essentially data frame objects and can thus be easily manipulated, e.g., using
functions from the `dplyr` package.

The following code shows how in this specific case, covariates relating to the
current treatment period ('PERIOD') and current treatment ('TREATMENT') are
sequentially derived and eventually used to create the 'FASTED' covariate.

Note that the 'EPOCH' field is not a standard field to be imported from the SDTM
data but can be included using the `keep` parameter in `add_administration()`:

```{r message = FALSE}
nif <- new_nif() %>% 
  add_administration(
    examplinib_fe, "EXAMPLINIB", analyte = "RS2023", keep = "EPOCH") %>% 
  add_observation(examplinib_fe, "pc", "RS2023") %>% 
  mutate(PERIOD = str_sub(EPOCH, -1, -1)) %>% 
  mutate(TREATMENT = str_sub(ACTARMCD, PERIOD, PERIOD)) %>% 
  mutate(FASTED = case_when(TREATMENT == "A" ~ 1, .default = 0)) 
```

These are again the first 3 lines of the resulting nif object:

```{r}
head(nif, 3)
```

# DATA EXPLORATION

## Data disposition

It is generally an excellent idea to explore data sets before proceeding into
more complex analyses. The `nif` package provides a host of functions to this
end. The following section shows some basic examples.

The `summary()` function generates a general overview on the data disposition in
a `nif` data set: 

```{r}
summary(nif)
```

Plotting the summary yields histograms of the baseline demographic covariates
and raw plots of the analytes over time. In the following code, ignore the
`ìnvisible(capture.output())` construct around the `plot()` function. Its sole
purpose is to omit some non-graphical output:

```{r warning=FALSE, results='hide', fig.width=3, fig.height=3}
invisible(capture.output(
  nif %>% 
    summary() %>% 
    plot()
))
```

## Plasma concentration data

`nif` objects can be easily plotted as time series charts using the generic
`plot()` function. While the output is a standard ggplot2 object that can be
further extended using ggplot2 functionality, the `plot()` function itself
includes extensive parameters to achieve the desired data visualization.

In its simplest form, `plot()` includes all analytes, and uses 'time after first
dose' ('TAFD') as the time metric:

```{r, warning=FALSE, fig.width=5, fig.asp=0.35}
plot(nif)
```

To check the integrity of the data set, it is often helpful to plot the analyte
concentrations over time-after-dose (TAD):

```{r warning=FALSE, fig.width=5, fig.asp=0.55}
nif %>% 
  plot(time="TAD", points=TRUE, lines=FALSE, log=TRUE)
```

To demonstrate the food effect on Cmax and Tmax on the individual level, the
below figure focuses on the first 24 hours on the linear scale and introduces
coloring based on the 'FASTED' covariate field:

```{r, warning=FALSE, fig.width=5, fig.asp=.7}
nif %>% 
  plot(color="FASTED", max_time=24, points=TRUE)
```

The following compares the mean plasma concentration profiles:

```{r, warning=FALSE, fig.width=5, fig.asp=.7}
nif %>% 
  plot(color="FASTED", max_time=24, mean=TRUE, points=TRUE)
```

Refer to the documentation of `plot.nif()` for further options.

## NIF viewer

`nif_viewer()` is a powerful exploratory tool that lets you interactively
explore all analyte profiles on an individual level. As the static nature of a
vignette does not allow to fully appreciate its potential, you are encouraged to
test `nif_viewer()` within your RStudio.

```{r, eval=FALSE, echo=TRUE}
nif_viewer(nif)
```





