---
title: "Imputations and assumptions used by make_nif()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputations and assumptions used by make_nif()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=F, message=F, warning=F}
library(nif)
```

This vignette outlines the inner workings of `make_nif()` to help understanding
what assumptions are made by the function and how e.g., time information is
derived.

## Baseline covariates

The subjects' height and body weight are included as standard fields in the 
NIF file. They are derived from the 'VS' domain, filtered by
`EPOCH=="SCREENING"`. If multiple measurements with this condition are included,
the mean value is used.

## Observations

### Specimen

The observations included in the NIF file are the pharmacokinetic concentrations
included in the 'PC' domain. Depending on the design of the underlying clinical
study, there may be data from different specimen types, e.g., blood, plasma,
urine, feces.
The specimen type to be included in the NIF file can be explicitly provided to
`make_nif()`, e.g., `make_nif(sdtm.data, spec="PLASMA")`. If not provided, the
most likely is used, i.e., either "PLASMA" or "BLOOD".

### Measurements not done

If the *permissible* field `PCSTAT` is used in the PC domain, providing
information on the status of the pharmacokinetic samples, data records with 
`PCSTAT=="NOT DONE"` are discarded.

### Observation time

The actual absolute date and time for each observation is taken from the `PCDTC`
field. The nominal (planned) time is taken from the `PCELTM` (planned elapsed
time) field. This *permissible* variable has the format "PTxxH" where *xx* is
the time since the last administration.

### Other fields required by NONMEM

For all observations, the compartment (`CMT`) is fixed to 2. Depending on the
further analysis, e.g., the structural model used in the nonlinear mixed-effects
modeling, the compartment may need to be changed, or different compartments for
the parent compound and metabolites may need to be defined.

The `DV` field in the NIF output is set to `PCSTRESN/1000`. This assumes that
the pharmacokinetic concentrations are given in ng/ml units. As per the NONMEM
conventions, `LNDV` is the natural log of `DV`. `MDV` (*missing dependent
variable*) is set to 1 for entries with missing `DV`.

For all observations, `EVID` is fixed to zero.






